Exploratory Analysis of Red Wine Quality by Jegan Kumar Ramanathan
==================================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r loading_packages}
# loading necessary packages
library(ggplot2)
library(GGally)
library(knitr)
```

```{r loading_data}
# loading the red wine quality dataset
redWine <- read.csv("wineQualityReds.csv")
```

I wouldn't call myself a wine connoisseur. But, I do like wines, especially, red
wines. This dataset consists of measurements of several chemical properties of 
over 1500 red wines. In addition to these input variables, we also have the
output variable named "quality" available for analysis. This output variable
is based on sensory data and it's a median of at least 3 evaluations made by 
wine experts. Objective here is to explore this data through effective 
descriptive statistics and visualizations to understand the relationships that 
exist between these variables, which inturn will help us in developing a 
statistical model to predict the quality of red wines.

More information on this dataset can be found [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt).

# Structure of the dataset

```{r structure_summary}
# removing row number column
redWine$X <- NULL

# reviewing the structure
str(redWine)
```

Note: the row number column has been removed as it is not useful for this 
analysis

# Univariate analysis

## Understanding the distribution of red wine quality

```{r quality}
ggplot(aes(x = quality), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('#36ae32')), 
                 binwidth = 1, stat = "count")
```

```{r quality_stats}
summary(redWine$quality)
```

The quality variable seems to be more or less normally distributed. But, the 
values are concentrated around the centre i.e. lot of the red wines have been 
rated as a 5 or 6 on a scale of 10. So, there are a lot of decent or average 
red wines in this dataset according to wine experts.

## Distribution of volatile acidity

```{r volatile_acidity}
ggplot(aes(x = volatile.acidity), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I("#e12301")), binwidth = 0.02) +
  scale_x_continuous(breaks = seq(0,1.6,0.2))
```

```{r volatile_acidity_stats}
summary(redWine$volatile.acidity)
```

Volatile acidity seems to be somewhat bi-modally distributed with a few outliers 
that have relatively high volatile acidity. High volatility in red wines results 
in an unpleasant, vinegar taste.

## Distribution of citric acid

```{r citric_acid}
ggplot(aes(x = citric.acid), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('yellow')), binwidth = 0.02) +
  scale_x_continuous(breaks = seq(0,1.1,0.1))
```
```{r citric_acid_stats}
summary(redWine$citric.acid)
```

Citric acid distribution is relatively uniform with two peaks at 0 and then at 
around 0.5. It does taper off on or after 0.5. Citric acid adds 'freshness' and 
flavor to wines.

## Distribution of residual sugar

```{r residual_sugar}
ggplot(aes(x = residual.sugar), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('gray')), binwidth = 0.2) +
  scale_x_continuous(breaks = seq(0,16,1))
```
```{r residual_sugar_stats}
summary(redWine$residual.sugar)
```

Residual sugar values are highly skewed and are right tailed. Residual sugar 
remaining after fermentation make the wines sweeter. Let's take a log tranform 
to get a different perspective on the distribution of this variable.

```{r log_residual_sugar}
ggplot(aes(x = residual.sugar), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('gray')), binwidth = 0.2) +
  scale_x_continuous(breaks = seq(0,16,1)) +
  scale_y_log10()
```

The distribution of the log10 transform of residual sugar further bolsters the 
observation that most of the red wines in this dataset have lower residual sugar 
content with a few exceptions.

## Distribution of free sulfur dioxide

```{r free_sulfur_dioxide}
ggplot(aes(x = free.sulfur.dioxide), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('orange')), binwidth = 1) +
  scale_x_continuous(breaks = seq(0,80,5))
```

```{r free_SO2_stats}
summary(redWine$free.sulfur.dioxide)
```

Free sulfur dioxide in high concentrations is increasingly noticeable through 
the smell and taste of wine. In this plot, we see a decreasing trend, as free 
sulfur dioxide increases, the count of red wines decreases. Also, there are a 
few outliers with relatively high amounts of free sulfur dioxide.

## Distribution of the ratio of free to total sulfur dioxide

```{r free_total_sulphur_dioxide}
ggplot(aes(x = free.sulfur.dioxide/ total.sulfur.dioxide), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('#c18617')), binwidth = 0.02)
```

Distribution of the ratio of free to total sulfur dioxide is normal but well 
distributed i.e. it has fatter tails.

## Distribution of pH

```{r pH}
ggplot(aes(x = pH), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('blue')), 
                 binwidth = 0.02) +
  scale_x_continuous(breaks = seq(2.7,4.1,0.1))
```

```{r ph_stats}
summary(redWine$pH)
```

pH values are normally distributed around the median/ mean of 3.31, indicating 
that most red wines are quite acidic.

## Distribution of sulphates

```{r sulphates}
ggplot(aes(x = sulphates), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('#f0c40b')), binwidth = 0.02) +
  scale_x_continuous(breaks = seq(0,2.4,0.2))
```
```{r sulphates_stats}
summary(redWine$sulphates)
```

Sulphates are additives used in red wines to increase sulfur dioxide gas levels 
which acts as an antimicrobial and antioxidant. This sulphate distribution is 
sort of skewed and normally distributed, with quite a few outliers.

## Distribution of alcohol

```{r alcohol}
ggplot(aes(x = alcohol), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('#f0680d')), binwidth = 0.1) +
  scale_x_continuous(breaks = seq(8,15,0.5))
```

```{r alcohol_stats}
summary(redWine$alcohol)
```

Alcohol percentage by volume seems to mostly range between 9 - 13. But, as 
alcohol % increases the count of red wines reduces i.e. they have a negative 
relationship. Like other variables, there are some outliers here too.

# Bivariate analysis

## Volatile acidity vs. Quality

```{r volatile_acidity_quality}
# changing quality data type from integer to factor
redWine$quality <- factor(redWine$quality)

ggplot(aes(x = quality, y = volatile.acidity), data = redWine) +
  geom_jitter(alpha = 0.6) +
  geom_boxplot(aes(fill = I("#f9ad93")), alpha = 1/2) +
  stat_summary(geom = "point", fun.y = mean, shape = 8, 
               size = 3, color = "red") +
  coord_cartesian(ylim = c(0.3,1.05)) +
  scale_y_continuous(breaks = seq(0.3,1.1,0.1))
```

As you can see clearly from the above box plots, as the quality increases the 
volatile acidity reduces. This makes sense as high volatile wines are considered
unpleasant.

Note: the quality variable was changed from integer to factor to facilitate the 
generation of certain types of plots

## Citric acid vs. Quality

```{r citric_acid_quality}
ggplot(aes(x = quality, y = citric.acid), data = redWine) +
  geom_jitter(alpha = 0.6) +
  geom_boxplot(aes(fill = I("yellow")), alpha = 1/2) +
  stat_summary(geom = "point", fun.y = mean, shape = 8, 
               size = 3, color = "blue") +
  coord_cartesian(ylim = c(0.0,0.55)) +
  scale_y_continuous(breaks = seq(0.0,0.55,0.1))
```

Unlike volatile acidity, as citric acid values increase so does the quality i.e. 
there seems to be a positive relationship between the amount of citric acid in 
red wines and its quality. This also is was expected as citric acids give 
'freshness' and flavor to wines which I would think is appreciated by the 
experts.

## Volatile vs. fixed acidity

```{r volatile_fixed_acidity}
# reloading data to convert quality back to an integer
redWine <- read.csv("wineQualityReds.csv")
# removing row number column
redWine$X <- NULL

ggplot(aes(x = fixed.acidity, y = volatile.acidity), data = redWine) +
  geom_point(alpha = 3/4) +
  stat_smooth(method = "lm", color = "brown")
```

```{r volatile_fixed_acidity_corr}
with(redWine, cor(x = quality, y = citric.acid))
```

As you can see from the scatter plot and correlation coefficient there is a 
negative relationship between volatile and fixed acidity, though the 
relationship is weak.

Note: the data was reloaded to change the quality variable data type to integer 
to facilitate generation of certain types of plots.

## Free vs. total sulfur dioxide

```{r free_total_SO2}
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), data = redWine) +
  geom_jitter(alpha = 1/2) +
  geom_smooth(method = "lm", color = "orange")
```

```{r free_total_SO2_corr}
with(redWine, cor(x = free.sulfur.dioxide, y = total.sulfur.dioxide))
```

It is obvious that there should be a relatively high correlation between free 
and total sulfur dioxide as they are dependent variables, total SO2 is a sum of 
free and bound SO2. The scatter plot and correlation coefficients just restate 
this obvious relationship.

## Computing correlations of all 11 input variables vs quality

```{r correlation}
for(i in c(1:11)){
  cat("Pearson's correlation coefficient of",names(redWine)[i],
      "and quality:", cor(redWine[,i], redWine$quality),"\n")
}
```

There is a medium strong negative correlation between volatile acidity & 
quality, and there is a weak positive correlation between citric acid and 
quality. This observation is in agreement with what we noticed in the scatter/ 
box plots before.

There is also a medium strong positive correlation between alcohol % and 
quality. So, it is worth investigating this relationship a bit further. Sulphate 
content also seems to have a weak positive correlation with quality which can be 
analyzed further as well.

# Multivariate analysis

## Volatile acidity vs. Quality by Alcohol %

```{r volatile_acidity_quality_alc}
ggplot(aes(x = quality, y = volatile.acidity), data = redWine) +
  geom_jitter(aes(color = alcohol), alpha = 3/4) +
  scale_colour_gradientn(colours = terrain.colors(10)) +
  geom_smooth(method = "lm", color = "red")
```

As you can see in the above plot, the alcohol content increases as the quality 
increases. This plot summarizes both the medium strong positive relationship of 
alcohol and quality, and the medium strong negative relationship of volatile 
acidity and quality.

## Free sulfur dioxide vs sulphates colored by quality

```{r free_SO2_sulphates_quality}
# changing quality data type from integer to factor
redWine$quality <- factor(redWine$quality)

ggplot(aes(x = sulphates, y = free.sulfur.dioxide), data = redWine) +
  geom_jitter(aes(color = quality), alpha = 3/4) +
  scale_color_brewer(palette = "RdYlGn") +
  geom_smooth(method = "lm")
```

As discussed before, sulphates are wine additives which increase SO2 gas levels. 
So, we would expect a correlation between free sulphur oxide (dissolved gas) and 
sulphates i.e. a positive relationship. But, in the above plot, there is a very 
weak positive correlation between these two variables. In addition to testing 
this relationship, this plot also shows how the quality of red wines vary with 
both sulphates and free sulfur dioxide. You could argue that as sulphates 
increase the quality increases (correlation coefficients calculated before 
suggest the same), but this is not the case with free sulfur dioxide. I guess 
what's happening here is that if the amount of free SO2 is low in certain wines, 
then just the right amount of sulphates are added to improve the wine.

Note: the quality variable was changed from integer to factor to facilitate the 
generation of certain types of plots

## pH vs fixed acidity colored by quality

```{r ph_fixed_acidity_quality}
ggplot(aes(x = fixed.acidity, y = pH), data = redWine) +
  geom_jitter(aes(color = quality), alpha = 0.65) +
  scale_color_brewer(palette = "Accent") +
  geom_smooth(method = "lm", color = "black")
```

Low pH values indicate an acidic solution, and high pH values indicate a basic 
solution. So, there is a direct relationship between pH and fixed acidity. As 
shown in the plot above, when fixed acidity increases, pH decreases. I added in 
quality as a third variable represented by the color to see whether there are 
any noticeable relationships between quality and pH and/or fixed acidity. It 
doesn't look like there is.

# Final Plots and Summary

```{r quality_final}
# reloading data to convert quality back to an integer to create scatter plots
redWine <- read.csv("wineQualityReds.csv")

# removing row number column
redWine$X <- NULL

ggplot(aes(x = quality), data = redWine) +
  geom_histogram(aes(color = I("black"), fill = I('#36ae32')), 
                 binwidth = 1, stat = "count") +
  labs(title = "Distribution of red wine quality as rated by experts") +
  ylab("Number of red wines") +
  xlab("Quality of red wines") +
  scale_y_continuous(breaks = seq(0,700,100)) +
  scale_x_continuous(breaks = seq(3,8,1))
```

This plot is a histogram of the quality variable which is a median of 3 or more 
red wine expert ratings. Understanding the distribution of this variable is a 
critical first step for analyzing the dataset and then subsequently in 
developing a good model that can effectively predict the quality of red wines 
not part of this dataset.

Note: the data was reloaded to change the quality variable data type to integer 
to facilitate generation of certain types of plots.

```{r citric_acid_quality_box_final}
# changing quality data type from integer to factor
redWine$quality <- factor(redWine$quality)

ggplot(aes(x = quality, y = citric.acid), data = redWine) +
  geom_jitter(alpha = 0.6) +
  geom_boxplot(aes(fill = I("yellow")), alpha = 1/2) +
  stat_summary(geom = "point", fun.y = mean, shape = 8, 
               size = 3, color = "blue") +
  coord_cartesian(ylim = c(0.0,0.55)) +
  scale_y_continuous(breaks = seq(0.0,0.55,0.1)) +
  ggtitle("Box plots of the amount of citric acid in red wines by quality") +
  xlab("Quality of red wines") +
  ylab("Amount of citric acid in g / dm^3")
```

These box plots clearly depict the positive relationship between citric acid 
and red wine quality. This could be used to develop a statistical model for 
red wine quality prediction.

Note: the quality variable was changed from integer to factor to facilitate the 
generation of certain types of plots

```{r volatile_acidity_quality_alc_f}
# reloading data to convert quality back to an integer to create scatter plots
redWine <- read.csv("wineQualityReds.csv")

# removing row number column
redWine$X <- NULL

ggplot(aes(x = quality, y = volatile.acidity), data = redWine) +
  geom_jitter(aes(color = alcohol), alpha = 3/4) +
  scale_colour_gradientn(colours = terrain.colors(10), 
                         name = "alcohol % by volume",
                         breaks = seq(8,15,2)) +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Scatter plot of amount of volatile acidity vs. quality, 
          colored by alcohol %") +
  ylab("Amount of volatile acidity in g / dm^3") +
  xlab(" Quality of red wines")
```

This colored scatter plot with straight line fitted through is an important 
visualization of how all three variables - volatile acidity, alcohol and quality 
are related to each other. I believe the understanding of these relationships 
will be crucial in developing a good model representative of red wine quality.

Note: the data was reloaded to change the quality variable data type to integer 
to facilitate generation of certain types of plots.

# Reflection

I explored all variables that I expected to have a reasonably strong 
relationship with red wine quality which would later help in developing a model 
for predicting the quality of other red wines not part of the dataset. 

I did struggle to generate multivariate plots as there were no categorical 
variables and just one discrete variable. I also took some time to wrap my head 
around what would be a good multivariate relationship to explore. 

Rest of the analysis was pretty smooth sailing. I was surprised by how alcohol 
was so strongly correlated with quality as I thought the amount of volatile 
acidity and citric acid were the key drivers of quality.

It would be good to develop a preliminary model for predicting wine quality with 
input variables as alcohol, volatile acidity, citric acid and sulphates, and 
test how the model performs. 

It would also be useful to transform or create new variables from some of these 
input variables and analyze whether these transformations produce a better 
relationship. 

Another area that can also be addressed are the outliers or anomalies found 
while analyzing these variables.
